---
title: "Arctic Fowl"
subtitle: "`polars` and `duckplyr`"
description: "Interpreted Languages can have nice things too!"
date: 07-04-2025
categories: [R, R-Package, Developer-Toolkit] 
citation: 
    url: https://asenetcky.dev/posts/2025-05-20-thinking-in-projects/
draft: true 
---

## What a Time to be Alive

In an era where AI is the all the rage, some of the most exciting things I have
encountered and have been playing around with lately have been remixes on old
favorites.  I'm talking of course about working with dataframes.  One of 
my favorites lately has been [polars](https://pola.rs/) and everytime
I'd use `polars` and then eventually do some work in R using the 
[tidyverse](https://www.tidyverse.org/), I would really feel some of the
niceities in `polars` that were missing in my day-to-day R workflow.

This is made all the more apparent to me because I adore working with
[dbplyr](https://dbplyr.tidyverse.org/) and its ability to lazily evaluate
R code as sql expressions and `collect()` the results into memory at a time
of my choosing. `polars` brings similar asbstractions to the table, however,
instead of a sql table being the target, dataframes are.  It's a genius 
idea that has really changed how I work and think about data and what I am
capable of doing with both resource rich, and starved computers before 
needing to scale up to other tools and techniques (if I even need to scale up
at all).

...

Enter [DuckDB](https://duckdb.org/),

I've always admired DuckDB from a distance. Dabbling here or there.  Some 
friends and colleagues of mine have thoroughly enjoyed it. I write a bunch
of SQL code for work, and for me, it has always felt like that - work. 
I believe that perhaps this has more to do with the locked down work 
environment with little to no modern tooling availible to anyone outside of IT 
(like Data Scientists). 

However, because of the harsh developer experience when dealing with SQL, 
I try to do  the absolute bare minimum I need to do and bail as fast as 
possible.  So DuckDB just never fell into my lap - until now.

[duckplyr](https://www.tidyverse.org/blog/2025/06/duckplyr-1-1-0/) joins
the `tidyverse` and now we have a fairly seamless drop-in replacement for
`dplyr` that leverages a `DuckDB` backend for everything it can, and falls
back to regular `dplyr` automatically when handed something it cannot
evaluate. For barely any setup at all - you get many of the same bells
and whistles of `polars`, but in quite _literally_ a tidy package.

Users get their lazy evaluation, their swapping between immediate and deferred
execution. Folks can work with data larger than the memory on their computer.
Like `polars` there are awesome functions for reading and writing data files.
If you're a tidyverse-native like myself, you can do all this without leaving
your comfortzone, and why would you want to be anywhere else?

In all seriousness, I think this is such an incredible win, `polars` syntax
was already fairly similar to `tidyverse` syntax in many ways.  There
was some cross-over with SQL verbage as well in the `tidyverse`, and DuckDB is
in-memory SQL and more.

In theory, I now have the ability to be dropped into basically
any team working with dataframes, and can just hit the ground running with 
_good_ options at my disposal regardless of the language of choice and the 
least amount of impact from context switching.

## Setup

You might mutter under your breath: "sure, that is all
great, but setup must be a pain" and I would argue no, or at least, it doesn't 
have to be.

Please refer to the documentation for the respective packages, especially
as they revolve over time.  However, setup might look something like this
for most users.

```{r}
#| label: actual-setup
#| echo: false

library(reticulate)
```


:::{.panel-tabset group="language"}

## R

With`renv` setup might look something like
this in the console:

```r
renv::install("duckplyr", "dplyr", "conflicted")
```

and then at the top of our R scripts:

```{r}
#| label: r-setup

suppressPackageStartupMessages({
    library(conflicted)
    library(duckplyr)
    conflict_prefer("filter", "dplyr") |> suppressMessages()
})

```

You _technically_ don't even need `conflicted`, but using that package
will provide a smoother user experience, and the explicitness in the will
help with reproducibility.

## Python
I'm using `uv` for python - so setup in the terminal would look 
_something_ like this:

```bash
$ mkdir my-project-folder
$ cd my-project-folder
$ uv init
$ uv add polars
```

and then at the top of our python scripts
(or bottom of your marimo notebooks ;p):

```{python}
#| label: python-setup

import polars as pl
```

:::

## Taking a Look

If it wasn't obvious before, I want to say this isn't going to be post where I 
pick a favorite and sell you on it. I really enjoy both packages and programming
languages.  

At the time of writing this, I am primarily an R programmer. I have spent more
time with python polars than anything duckdb-related, and quite a bit of time
recently with polars. 

Still, all of the time with python has been considerably 
less time compared to R. I have experimented a little bit with the duckdb cli
tool in the past, but nothing more than an afternoon or two.  This is my first
real foray into working with the `duckplyr` package.  All this to say, that
I am not an expert in either of these tools and I am likely not working
with them in the most idiomatic way. I'm mostly just here to crush
parquet files and have a good time.

`duckplyr` has some example data bundled with the package. We'll work with that.
First we'll save to parquet file into a temporary folder. In the real world 


::: {.callout-note}
Ideally, users would just use the data in memory and pass it around between 
the languages using something like `arrow` and `reticulate` to handle 
zero-copy streaming etc...

However, I think using local files more clearly displays the pros of these
packages - so that is the route we're going to take.
:::

### Example Data
```{r}
#| label: save-example-data

library(fs)

my_temp_dir <- fs::file_temp() |> fs::dir_create()
data_path <- fs::path(my_temp_dir, "temp-flight-data", ext = "parquet")

# save path to python object
py_data_path <- reticulate::r_to_py(data_path)

duckplyr::flights_df() |> duckplyr::compute_parquet(data_path) |> invisible()
```

These data aren't large by any means, but they are readily available - and 
`duckplyr` uses these data in their documentation. I was able to use 7GB+ 
parquet files on my humble laptop working with the same code below without
crashing my IDE, which is a modern miracle.

### Nanoparquet

I want to take a quick aside and mention that `nanoparquet` has some nice tools 
for inspecting parquets without reading the entire file into memory as well.
[nanoparquet](https://nanoparquet.r-lib.org/) is completely dependency free
and a worthy companion to `duckplyr`.

Since we have a parquet file - let's take a quick peek, and then go back to
using the `duckplyr` and `polars`.

```{r}
#| label: nanoparquet

library(nanoparquet)

# grab very basic info about the file and data structure
info <- nanoparquet::read_parquet_info(data_path)
```

```r
# grab much more granular information about the data schema
schema <- nanoparquet::read_parquet_schema(data_path)

# grab the info above and more!
metadata <- nanoparquet::read_parquet_metadata(data_path)
```

```{r}

# scope out the basic info
glimpse(info)
```


We have the parquet, we know the data's basic dimensions, and we know that
the number rows probably won't blow up our IDE session. Let's dive in, but
we'll pretend there are a few more 0's after that `info$num_rows` and show off
how we might dance around that problem.

## Diving Right In

We're going to look at some common actions side by side.

:::{.panel-tabset group="language"}

## R

```{r}
#| label: r-read-parquet

ducks <- 
    read_parquet_duckdb(
        data_path, 
        prudence = "stingy"
        )

# no collect - no row count
head(ducks)

# collect -  notice the row count
ducks |> 
    head() |> 
    arrange() |>
    collect()


ducks |> 
    head() |> 
    collect() |> 
    glimpse()

# basic group by
ducks |> 
    # notice the use of `.by` in summarise and not group_by()
    summarise(
        count = n(),
        .by = month
    ) |> 
    head(n=5) |> 
    arrange(month) |> 
    collect()
```


## Python


```{python}
#| label: py-read-parquet

bears = pl.scan_parquet(r.py_data_path)

# collect into a dataframe
bears.head().collect()

bears.head().collect().glimpse()

(
    bears.group_by(pl.col("month"))
    .len(name="count")
    .sort("month")
    .head(5)
    .collect()
)

```

:::


## Taking a step back

Killer features: 

### Lazy Evaluation

`polars` and `duckplyr` are both able to lazily evaluate data transformations.

Users can take objects in-memory, or a file, and queue up actions to perform
later on.  Why is this important? This is primarily a plus when working with
large amounts of data that you could not possibly store into the RAM of your
computer, or it would be prohibitively expensive to provision a cloud computer
with the amount of RAM needed to. 

What if our example data was a 7GB parquet file and it was 259 million
rows long. I couldn't possibly hope to read that whole file in _and then_
start filtering and grouping and other data wrangling. Lazy evaluation
allows users to effectively stream the data from memory or a file, taking
only the needed bits, doing only the necessary operations, and doing nothing
more.

Delaying the execution of actions also allows for the packages to implement
query optimization, where the user describes the steps and what they want to see
and the package will work out the best steps to get there.

### So Fast

TODO: duckplyr prudence
lavish - "think eager" always materialize
stingy - avoid spending at all cost 
thrifty - use memory wisely

polars lazyframe -lazy and dataframe eager.

TODO: get the dimensions from the pkgs



If I tried to `collect()` even one of these - at least on this computer,
I run out of memory everytime. However, I can have `ducks` and `bears`
living together in harmony!


```{r}

# I can't glimpse and collect()
# but I can head and collect and glimpse
ducks |>
    head() |> 
    select(1:10) |> 
    collect() |> 
    glimpse()

```



```{python}
bears.head().collect().glimpse()

```
