---
title: "Arctic Fowl"
subtitle: "`polars` and `duckplyr`"
description: "Interpreted Languages can have nice things too!"
date: 07-04-2025
categories: [R, R-Package, Developer-Toolkit] 
citation: 
    url: https://asenetcky.dev/posts/2025-05-20-thinking-in-projects/
draft: true 
---

## What a Time to be Alive

In an era where AI is the all the rage, some of the most exciting things I have
encountered and have been playing around with lately have been remixes on old
favorites.  I'm talking of course about working with dataframes.  One of 
my favorites lately has been [polars](https://pola.rs/) and everytime
I'd use `polars` and then eventually do some work in R using the 
[tidyverse](https://www.tidyverse.org/), I would really feel some of the
niceities in `polars` that were missing in my day-to-day R workflow.

This is made all the more apparent to me because I adore working with
[dbplyr](https://dbplyr.tidyverse.org/) and its ability to lazily evaluate
R code as sql expressions and `collect()` the results into memory at a time
of my choosing. `polars` brings similar asbstractions to the table, however,
instead of a sql table being the target, dataframes are.  It's a genius 
idea that has really changed how I work and think about data and what I am
capable of doing with both resource rich, and starved computers before 
needing to scale up to other tools and techniques (if I even need to scale up
at all).

...

Enter [DuckDB](https://duckdb.org/),

I've always admired DuckDB from a distance. Dabbling here or there.  Some 
friends and colleagues of mine have thoroughly enjoyed it. I write a bunch
of SQL code for work, and for me, it has always felt like that - work. 
I believe that perhaps this has more to do with the locked down work 
environment with little to no modern tooling availible to anyone outside of IT 
(like Data Scientists). 

However, because of the harsh developer experience when dealing with SQL, 
I try to do  the absolute bare minimum I need to do and bail as fast as 
possible.  So DuckDB just never fell into my lap - until now.

[duckplyr](https://www.tidyverse.org/blog/2025/06/duckplyr-1-1-0/) joins
the `tidyverse` and now we have a fairly seamless drop-in replacement for
`dplyr` that leverages a `DuckDB` backend for everything it can, and falls
back to regular `dplyr` automatically when handed something it cannot
evaluate. For barely any setup at all - you get many of the same bells
and whistles of `polars`, but in quite _literally_ a tidy package.

Users get their lazy evaluation, their swapping between immediate and deferred
execution. Folks can work with data larger than the memory on their computer.
Like `polars` there are awesome functions for reading and writing data files.
If you're a tidyverse-native like myself, you can do all this without leaving
your comfortzone, and why would you want to be anywhere else?

In all seriousness, I think this is such an incredible win, `polars` syntax
was already fairly similar to `tidyverse` syntax in many ways.  There
was some cross-over with SQL verbage as well in the `tidyverse`, and DuckDB is
in-memory SQL and more.

In theory, I now have the ability to be dropped into basically
any team working with dataframes, and can just hit the ground running with 
_good_ options at my disposal regardless of the language of choice and the 
least amount of impact from context switching.

## Setup

You might mutter under your breath: "sure, that is all
great, but setup must be a pain" and I would argue no, or at least, it doesn't 
have to be.

Please refer to the documentation for the respective packages, especially
as they revolve over time.  However, setup might look something like this
for most users.

```{r}
#| label: actual-setup
#| echo: false

library(reticulate)
```


:::{.panel-tabset group="language"}

## R

I'm using `renv` so setup might initially look something like
this in the console:

```r
renv::install("duckplyr", "dplyr", "conflicted")
```

and then at the top of our R scripts:

```{r}
#| label: r-setup

suppressPackageStartupMessages({
    library(conflicted)
    library(duckplyr)
    conflict_prefer("filter", "dplyr") |> suppressMessages()
})

```

## Python
I'm using `uv` for python - so setup in the terminal would look 
_something_ like this:

```bash
$ uv init
$ uv add polars
```

and then at the top of our python scripts:

```{python}
#| label: python-setup

import polars as pl
```

:::

## Taking a Look

If it wasn't obvious before, I want to say this isn't going to be post where I 
pick a favorite and sell you on it. I really enjoy both packages and programming
languages.  

At the time of writing this, I am primarily an R programmer. I have spent more
time with python polars than anything duckdb-related, and quite a bit of time
recently with polars. 

Still, all of the time with python has been considerably 
less time compared to R. I have experimented a little bit with the duckdb cli
tool in the past, but nothing more than an afternoon or two.  This is my first
real foray into working with the `duckplyr` package.  All this to say, that
I am not an expert in either of these tools and I am likely not working
with them in the most idiomatic way. I'm mostly just here to crush
parquet files and have a good time.

`duckplyr` has some example data bundled with the package. We'll work with that.

```{r}
library(fs)

my_temp_dir <- fs::file_temp() |> fs::dir_create()
data_path <- fs::path(my_temp_dir, "temp-flight-data", ext = "parquet")

duckplyr::flights_df() |> duckplyr::compute_parquet(data_path) |> invisible()
```


I want to take a quick aside and mention that `nanoparquet` has some nice tools 
for inspecting parquets without reading the entire file into memory.
Since we have a parquet file - these functions would be the ideal choice for
taking a quick look. We'll take a quick peek, and then go back to using the
packages. 

```{r}
library(nanoparquet)

info <- nanoparquet::read_parquet_info("flights.parquet")
schema <- nanoparquet::read_parquet_schema("flights.parquet")
metadata <- nanoparquet::read_parquet_metadata("flights.parquet")


# dimensions and file info
head(info)

# variable, variable types and conversion metrics
head(schema)

# everything above and the kitchen sink
# here is a sample of the new stuff
metadata |> 
    purrr::discard_at(c("file_meta_data", "schema")) |> 
    purrr::map(head, 1)
```


Now we have are parquet - let's dive in.

Polars and duckplyr are both able to lazily evaluate
data.

TODO: duckplyr prudence
lavish - "think eager" always materialize
stingy - avoid spending at all cost 
thrifty - use memory wisely

polars lazyframe -lazy and dataframe eager.

TODO: get the dimensions from the pkgs

:::{.panel-tabset group="language"}

## R

```{r}
# reading parquet - lazily

ducks <- read_parquet_duckdb("flights.parquet", prudence = "stingy")

# no collect
ducks |> head()

# collect -  notice the rows
ducks |> head() |> collect()


```

## Python

```{python}
bears = pl.scan_parquet("flights.parquet")

# lazyframes - won't print any data without collection

# collect into a dataframe
bears.head().collect()
```

:::


If I tried to `collect()` even one of these - at least on this computer,
I run out of memory everytime. However, I can have `ducks` and `bears`
living together in harmony!


```{r}

# I can't glimpse and collect()
# but I can head and collect and glimpse
ducks |>
    head() |> 
    select(1:10) |> 
    collect() |> 
    glimpse()

```


St

```{python}
bears.head().collect().glimpse()

```
